---
title: "North Carolina_County_Level_Poverty_Analysis"
author: "Waleed Abdulla"
date: "2025-04-04"
output: pdf_document
---

**1 Data Preparation**

Note that the following libraries are uploaded in the background to create a cleaner markdown file: knitr, fpp3, eeptools, usmap, stringr, urca
```{r setup, include=FALSE}
library(knitr)
library(fpp3)
library(eeptools)
library(usmap)
library(stringr)
library(urca)
```

**1.1 SAIPE data**

```{r, warning=FALSE}
NCRaw <- read.csv("Data/SAIPE_04-04-2025.csv") %>% 
  select(Year, ID, Name, Poverty.Universe, Number.in.Poverty) %>% 
  rename(fips = ID, `County Population` = Poverty.Universe, `Number in Poverty` = Number.in.Poverty)

#decomma() found in eeptools library transforms string numbers with commas into numeric values
NC <- NCRaw %>% mutate(`County Population` = decomma(`County Population`),
                       `Number in Poverty` = decomma(`Number in Poverty`)) %>% na.omit()
```

This project's state name, abbreviation, and FIPS code is North Carolina, NC, and 37, respectively.

```{r}
NC %>% select(Name) %>% unique() %>% count()
NC %>% group_by(Name) %>% summarize(Max = max(`County Population`)) %>% arrange(desc(Max)) %>% head(5)
```
There are 100 counties.
The largest county ever recorded was Wake County with 1,173,780 people.

```{r}
NC %>% filter(Year == 2023) %>% arrange(desc(`County Population`)) %>% head(9)
topNine <- NC %>% filter(Year == 2023) %>% arrange(desc(`County Population`)) %>% head(9) %>% pull(Name)
```
The above are the nine largest counties by the most recent year in the dataset (2023).

```{r}
NCMap <- NC %>% filter(Year == 2023) %>% mutate(values = `County Population`, fips = fips) %>% select(fips, values)
plot_usmap(regions = "counties", include = "North Carolina", data = NCMap) +
  scale_fill_viridis_c(option = "A", direction = -1) + labs(fill = "Population") + theme(legend.position = "left")

NCGIS <- NC %>% filter(Year == 2023)
write.csv(NCGIS, "Data/North_Carolina_Counties_and_Poverty.csv")
```

```{r}
NC9 <- NC %>% filter(Name %in% topNine)
NC9 %>% ggplot(aes(x=Year,y=`Number in Poverty`,colour = Name)) + geom_line()
```

**1.2 County SNAP Benefits**

```{r}
SNAPRaw <- read.csv("Data/cntysnap.csv", skip=5) %>% 
  filter(State.FIPS.code == 37, County.FIPS.code != 0) %>% 
  mutate(fips = paste(as.character(State.FIPS.code), sprintf("%03d", County.FIPS.code), sep = "") %>% as.integer()) %>%
  select(-c(County.FIPS.code, State.FIPS.code))

Snap <- SNAPRaw %>% pivot_longer(!c(Name,fips), names_to = "Year", values_to = "Snap") %>%
  mutate(Year = str_remove(Year,"Jul.") %>% as.integer(), 
         Name = str_remove(Name,", NC"), Snap = decomma(Snap) %>% as.integer())

Snap %>% head(5)

Snap9 <- Snap %>% filter(Name %in% topNine)
Snap9 %>% ggplot(aes(x=Year,y=Snap,colour = Name)) + geom_line()
```

**1.3 State IRS Data**

```{r}
IrsRaw <- read.csv("Data/irs.csv",skip=5)

Irs <- IrsRaw %>% filter(State.FIPS.code == 37) %>% mutate(Poor.exemptions = decomma(Poor.exemptions), fips = State.FIPS.code) %>%
  select(fips, Poor.exemptions, Year)
Irs %>% ggplot(aes(x=Year,y=Poor.exemptions)) + geom_line()
```

**1.4 Merging the data**

```{r, warning=FALSE}
NCaroR <- left_join(Snap,Irs,by="Year")
NCaroR <- left_join(NCaroR,NC)
NCaro <- NCaroR %>% mutate(fips = fips.x) %>% select(-c(fips.y,fips.x)) %>% 
  filter(Year > 1997) %>% select(Year, fips, Name, `Number in Poverty`, `County Population`, Snap, Poor.exemptions)
NCaro %>% head()
```

Now check for any problems

```{r}
NCaro$fips %>% unique()
NCaro %>% group_by(fips) %>% summarize(count = n()) %>% filter(count != 25)
```
All 100 counties have 25 rows, from 2023 to 1998. There are only 100 unique counties in the dataset too.

```{r}
NCaro %>% is.na() %>% mean()
```

There are no NA values (removed in beginning).

After checking the county changes website, only two North Carolina Counties had substantial boundary changes where, in the 1990s, Carteret County received five square miles and lost 16 square miles from Craven County.

```{r,warning=FALSE}
NCTS <- NCaro %>% as_tsibble(Year, key=c(fips,Name))
NCTS %>% head()
NCTS %>% ggplot(aes(x=log(Snap),y=log(`Number in Poverty`), color = Name)) + geom_point(show.legend = FALSE)
NCTS %>% ggplot(aes(x=Poor.exemptions,y=`Number in Poverty`, color = Name)) + geom_point(show.legend = FALSE)
NCTS %>% ggplot(aes(x=log(`County Population`),y=log(`Number in Poverty`), color = Name)) + geom_point(show.legend = FALSE)
```


-------------

**Part 2: Linear Models**

-------------


**2.1 Variable Selection**

```{r}
fitSet <- NCTS %>%
  model(
    P = TSLM(log(`Number in Poverty`) ~ log(`County Population`)),
    PS = TSLM(log(`Number in Poverty`) ~ log(`County Population`) + log(Snap)),
    PSE = TSLM(log(`Number in Poverty`) ~ log(`County Population`) + log(Snap) + log(Poor.exemptions)),
    S = TSLM(log(`Number in Poverty`) ~ log(Snap)),
    SE = TSLM(log(`Number in Poverty`) ~ log(Snap) + log(Poor.exemptions)),
    E = TSLM(log(`Number in Poverty`) ~ log(Poor.exemptions)),
    ES = TSLM(log(`Number in Poverty`) ~ log(`County Population`) + log(Poor.exemptions))
    )

glance(fitSet) %>% group_by(.model) %>% summarize(AIC = mean(AIC))
```

Based on the AIC test, the model that uses the Population and Snap variables does the best across all counties by a small margin.

```{r}
NCTS %>% filter(Name %in% topNine) %>% autoplot(`Number in Poverty`, color = "black") + 
  facet_wrap("Name", scales = "free_y") + 
  geom_line(data = fitSet %>% fitted() %>% filter(.model == "PS", Name %in% topNine), aes(y = .fitted), color = "red")
```

**2.2 Residual Analysis**

```{r}
fitSet %>% filter(Name %in% topNine) %>% select(PS) %>% augment() %>% autoplot(.innov)
```

```{r}
fitSet %>% filter(Name %in% topNine) %>% select(PS) %>% augment() %>%
  features(.innov, ljung_box, lag = 10) %>% filter(lb_pvalue <= 0.05)
```
Two counties' residuals are significantly different from white noise.

The linear model does a good job of predicting the number in poverty since only two counties are significantly different from white noise.


-------------

**Part 3: Stochastic Models**

-------------


**3.1 Single County Forecasts**

```{r}
WC <- NCTS %>% filter(Name == topNine[1]) #Wake County

fitSetWC <- WC %>% model(NAIVE = NAIVE(log(`Number in Poverty`)),
                         MEAN = MEAN(log(`Number in Poverty`)),
                         SES = ETS(log(`Number in Poverty`) ~ error("A") + trend("N") + season("N")),
                         HoltD = ETS(log(`Number in Poverty`) ~ error("A") + trend("Ad") + season("N")),
                         HoltND = ETS(log(`Number in Poverty`) ~ error("A") + trend("A") + season("N")),
                         ARIMA = ARIMA(log(`Number in Poverty`)))

fitSetWC %>% forecast(h = "5 years") %>% autoplot(WC) + facet_wrap(".model")

fitSetWC %>% accuracy() %>% group_by(.model) %>% summarize(RMSE = mean(RMSE))
```

Based on the RMSE scoring, the Holt’s additive linear damped method smoothing is the best for this county.

**3.2 Exponential Smoothing Models**

```{r}
fitSetES <- NCTS %>% 
  model(SES = ETS(log(`Number in Poverty`) ~ error("A") + trend("N") + season("N")),
        HoltD = ETS(log(`Number in Poverty`) ~ error("A") + trend("Ad") + season("N")),
        HoltND = ETS(log(`Number in Poverty`) ~ error("A") + trend("A") + season("N"))
        )

fitSetES %>% glance() %>% group_by(.model) %>% summarize(AICc = mean(AICc))
```

Based on the AICc score, the simple exponential smoothing model is best since it has the lowest AICc.

**3.3 ARIMA Models**

```{r}
ARIMAs <- NCTS %>% model(AA = ARIMA(log(`Number in Poverty`)))
ARIMAs %>% mutate(Type = ARIMAs$AA %>% as.character()) %>% group_by(Type) %>% summarise(Amount = n())
```

The most commonly selected models are ARIMA(1,0,0) w/ mean and ARIMA(0,1,0) (random walk).

```{r}
fitSetA <- NCTS %>% model(A010 = ARIMA(log(`Number in Poverty`) ~ pdq(0,1,0)),
                          A100M = ARIMA(log(`Number in Poverty`) ~ pdq(1,0,0) + 1))

fitSetA %>% glance() %>% group_by(.model) %>% summarize(AICc = mean(AICc))
```

Based on the AICc evaluator, the ARIMA(0,1,0) (random walk) model is best since it has the lowest AICc. 

**When Including $+ 1$ to the ARIMA(1,0,0) with mean model, the results did not change. After some research I found that pdq(1,0,0) internally includes the + 1 intercept.**

**3.4 Cross Validation**

```{r, warning=FALSE}
finalfit <- NCTS %>% stretch_tsibble(.init = 20) %>% model(
  SES = ETS(log(`Number in Poverty`) ~ error("A") + trend("N") + season("N")),
  A010 = ARIMA(log(`Number in Poverty`) ~ pdq(0,1,0)))

fff <- finalfit %>% forecast(h = "5 years") 

fff %>% accuracy(NCTS) %>% select(.model, fips, Name, .type, RMSE) %>% group_by(.model) %>% summarize(RMSE = sum(RMSE))
```

Based on the summed RMSE, the Simple Exponential Smoothing model preformed the best on cross validation.


-------------

**Part 4: Forecasts**

-------------


```{r, cache=TRUE}
bestFit <- NCTS %>% model(ETS(log(`Number in Poverty`) ~ error("A") + trend("N") + season("N")))

bff <- bestFit %>% forecast(h = "5 years")

prediction <- bff %>% filter(Year == 2027) %>% pull(.mean)

NorthCarolina <- NCTS %>% filter(Year == 2022) %>% 
  mutate(Prediction = prediction, 
         "Percentage Increase" = (prediction - `Number in Poverty`)/`County Population`,
         "Poverty Increase" = (Prediction - `Number in Poverty`)) %>%
  select(c(Year, fips, Name, `Number in Poverty`, Prediction, `Poverty Increase`, `Percentage Increase`)) %>%
  rename(values = `Poverty Increase`)

plot_usmap(regions="counties",include="North Carolina", data = NorthCarolina, values = "Percentage Increase") +
  labs(fill = "Predicted Poverty Increase") + scale_fill_viridis_c(option = "B", direction = -1) + theme(legend.position = "left")

NorthCarolina %>% rename(`Poverty Increase` = values) %>% arrange(desc(`Percentage Increase`)) %>% 
  select(-`Poverty Increase`) %>% head(5)
#Note that removing the `Poverty Increase` column is purely for visual purposes.

NC5 <- NorthCarolina %>% arrange(desc(`Percentage Increase`)) %>% head(5) %>% pull(Name)

bff %>% filter(Name %in% NC5) %>% autoplot(NCTS, colour = "red") + facet_wrap("Name")

write.csv(NorthCarolina, "Data/North_Carolina_Predicted_Poverty_Increase.csv")
```


We predict the that Halifax, Bladen, Northampton, Richmond, and Robeson County are the five counties that will have the highest percentage increase in poverty over the next five years.